ODGI: understanding pangenome graphs

### 2021-10-27 manuscript/main.tex
remove or replace all footnotes

\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1\footnote{\url{https://github.com/GFA-spec/GFA-spec/blob/master/GFA1.md} (accessed Oct 2021)}  format. It can reveal a graph’s internal structures for e.g. debugging processes.

@@

%For the establishment of a layout in two dimensions \textit{odgi layout} applies a 2D adjusted PG-SGD algorithm. A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}.

@@

Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. It outputs pangenome statistics in TSV or YAML textual file formats.
MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
ODGI also offers more advanced tools for the interrogation of the graphs. \textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. It provides tab delimited or JSON `bins'.
This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. \textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
Collapsed sequences with highly identical repeats indicate intricate loci.
Complex motifs can be also detected with \textit{odgi degree}, which returns the node degree as defined by query criteria.
High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
In Fig. \ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.

>>>>

\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 \citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.

@@

%For the establishment of a layout in two dimensions \textit{odgi layout} applies a 2D adjusted PG-SGD algorithm. A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\{\url{https://github.com/chfi/gfaestus}}.

@@

Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. It outputs pangenome statistics in TSV or YAML textual file formats.
MultiQC's \citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} 
provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
ODGI also offers more advanced tools for the interrogation of the graphs. \textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. It provides tab delimited or JSON `bins'.
%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion
\textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
Collapsed sequences with highly identical repeats indicate intricate loci.
Complex motifs can be also detected with \textit{odgi degree}, which returns the node degree as defined by query criteria.
High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
In Fig. \ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
